Index measures for oak decline severity using phenotypic descriptors

### 2019-12-17 manuscript/manuscript.Rmd
finished machine learning discussions section

---
title: "Index measures for oak decline severity using phenotypic descriptors"
output: 
  #bookdown::word_document2:
  # toc: false
  bookdown::pdf_document2: 
    toc: false
    number_sections: false
bibliography: references.bib
---

@@

Diameter at breast height (DBH) was collected at a stem height of 1.3m, to the nearest 0.1cm.
Total height, timber height and lowest live crown were measured using an inclinometer to the nearest 0.1m
The timber height was measured at either 7cm top stem diameter or where no main stem could be distinguished.
For the crown, crown contact, missing crown and absolute crown density percentages were visually estimated within 5% classes

@@

The interactions of key raw descriptors identified from the overall importance scores and their associations with the decline indexes was investigated by using simulated descriptor values to construct index model responses surfaces for specific cases.
This approach was similar to response surface methodology without the need for probabilistic approximation due to the low computation time required for index calculation (ref).

@@

The DAI is a much more complex index with more descriptors contributing to it than the PDI such as bleed prevalence, DBH and crown condition descriptors. 
Simulated response surfaces for the DAI are shown in Figure 4. 
These response surfaces show circumstances under which bleed prevalence related descriptors (Figures 4a, b & c) and crown condition descriptors (Figures 4d, e & f) change for non-symptomatic AOD or COD trees (Figures 4a & d respectively), moderate AOD or COD (Figures 4b & e respectively) and severe AOD or COD (Figures 4c & f respectively). 
Increasing active bleed, staining size and frequency contributed to greater positive values of the DAI and more severe AOD. 
Also higher DAI values could be obtained from trees with greater stature (Figure 4c). 
Crown transparency and missing crown have no influence on the DAI for trees of neutral decline type (Figure 4d).
Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures 4e & f) and therefore more severe COD.
The relationship of relative tree stature and the acuteness of the decline likely reflects the period of time under which these trees have been afflicted. 
AOD occurs over a much shorter period compared COD and therefore has less time in which to inhibit growth of the tree as a whole.

@@

The phenotypic descriptors were selected for use in this study purposefully

- Suitability of collected phenotypic descriptors and decline indexes for describing decline status
- Index thresholds for category definitions (healthy, declining, severe decline etc.)
- Validation of decline scores - chicken and paradox
- Direct validation difficult 
- Refined definition of remission - repeat measurement
- With different data, it is likely that the MDS dimensions may require inversion prior to min-max scaling to ensure their correct orientation.

@@

These two indexes provide a two dimensional continuum under which the spectrum of decline severity and type can be assessed.

The two dimensional space provided by these phenotypic indexes, for measuring decline severity is shown in Figure 5.

@@

- reasons for selecting machine learning approach
- use of unsupervised over supervised analysis
- Semi-supervised can be used if unsupervised model is unable to pull out decline severity scale
- For given context, descriptors not contributing to the indexes could be removed
- Concession of model interpretability to predictive capability by using a machine learning approach

Machine learning approaches are usually seen as black box 
- seen as black box but LIME and response surfaces enable interpretation and therefore trust
- LIME can be applied to out of place cases
- Important descriptors identified model-wide aren't necessarily relevant to localised cases.

- use of indexes derived from machine learning are context agnostic

### Applications and wider scalability

@@

The current context in which these decline indexes are presented is based on a very focus set of descriptors on sites with a high proportion of declining trees.

Scaling this technique to enable generalised naive computation of decline indexes for any tree sampled in the UK would require a substantially larger training data set that incorporates the descriptive diversity of oak decline in the population as whole (ref).
Scaling to this level would also require an appropriate method of tree size correction that could be applied on an individual tree basis.
A potential solution to this would be to correct trees based on estimated age from DBH as proposed by @white1998.
This would enable the development of web based tools that can then provide access to

- Discussion of applications to forestry monitoring and management
- Refinement of protocol for large scale surveying

@@

@@

- Crown condition descriptors most important for describing overall decline severity in the PDI.
- Bacterial bleed related descriptors are most important for describing decline acuteness.
- Decline indexes provide intuitive measures of decline severity based on simple phenotypic measurements that do not require specialist equipment
- Encouraged to apply this approach in other contexts

The decline indexes presented here provided intuitive measures 
The approach presented here is easily adaptable to incorporate any phenotype

>>>>

---
title: "Index measures for oak decline severity using phenotypic descriptors"
output: 
  word_document: default
  # pdf_document: default
bibliography: references.bib
---

@@

The phenotypic descriptors collected were based on @innes1990.
Diameter at breast height (DBH) was collected at a stem height of 1.3m, to the nearest 0.1cm.
Total height, timber height and lowest live crown were measured using an inclinometer to the nearest 0.1m
The timber height was measured at either 7cm top stem diameter or where no main stem could be distinguished.
For the crown, crown contact, missing crown and absolute crown density percentages were visually estimated within 5% classes

@@

The interactions of key raw descriptors identified from the overall importance scores and their associations with the decline indexes was investigated by using simulated descriptor values to construct index model response surfaces for specific cases.
This approach was similar to response surface methodology without the need for probabilistic approximation due to the low computation time required for index calculation [@abbasi2012].

@@

The DAI is a much more complex index with more descriptors contributing to it than the PDI such as bleed prevalence, DBH and crown condition descriptors. 
Simulated response surfaces for the DAI are shown in Figure 4. 
These response surfaces show circumstances under which bleed prevalence related descriptors (Figures 4a, b & c) and crown condition descriptors (Figures 4d, e & f) change for non-symptomatic AOD or COD trees (Figures 4a & d respectively), moderate AOD or COD (Figures 4b & e respectively) and severe AOD or COD (Figures 4c & f respectively). 
Increasing active bleed, staining size and frequency contributed to greater positive values of the DAI and more severe AOD. 
Also higher DAI values could be obtained from trees with greater stature (Figure 4c). 
Crown transparency and missing crown have no influence on the DAI for trees of neutral decline type (Figure 4d).
Similarly to the PDI, trees with smaller stature and poorer crown condition give lower DAI values (Figures 4e & f) and therefore more severe COD.

@@

The decline indexes presented here show 
These two indexes provide a two dimensional continuum under which the spectrum of decline severity and type can be assessed (Figure 5).

- Indexes place the descriptors into single intuitive scores based on a wide number of factors known to be heavily influenced by oak decline severity
- Suitability of collected phenotypic descriptors and decline indexes for describing decline status

@@

- Index thresholds for category definitions (healthy, declining, severe decline etc.)

- Association of PDI and overall oak decline with crown condition well documented
- Association of crown condition and AOD severity. Could not previously be shown by @denman2014.
- Bleed prevalence and AOD
- Agrillus exit hole occurance with stem bleeding - smaller area surveyed compared to bleeding. Trees with bleeds do not always have stem bleeds [@brown2017;@vansteenkiste2004]
- Size differences between AOD and COD trees

The relationship of relative tree stature and the acuteness of the decline likely reflects the period of time under which these trees have been afflict. 
AOD occurs over a much shorter period compared COD and therefore has less time in which to inhibit growth of the tree as a whole.

- Refined definition of remission - repeat measurement

@@

An unsupervised approach was favoured in this context due to the concession that there know difficulties and inaccuracy of manual field decline status classification (ref).
This is highlighted in Figure 1 where a numerous trees would be considered to have been misclassified as they fall into regions of the decline indexes associated with a different decline status (eg. COD trees with DAI values > 0.5).
The unsupervised approach has worked in this context in aptly describing decline severity as the phenotypic descriptors selected were highly specific and well suited to describing the decline severity problem.
In situations were descriptors may well be less suited, perhaps as a result of technical limitations, the use of a semi-supervised approach may be favourable (ref).
The unsupervised approach to derive the decline indexes would be applied to a subset of the observations on which the descriptors are known to perform well and that would be considered benchmark observations by the investigators.
Indexes could then be attained for the rest of the observations by prediction based on supervised models of the benchmark subset (ref).
This approach would be more suitable than a fully supervised approach as it would limit the inaccuacies that would be introduced by manual classification and still allow the data driving flexibility of the unsupervised approach.

Machine learning approaches are usually seen as black box and a lack of model interpretability is a concession in this approach however the are substantial gains in terms of predictive capability for model application in real world situations (ref).
The assessment descriptor contributions to the index models are an integral part of this approach.
Inout of knowledge of the biological problem by the investigator at this stage is essential in ensuring that the models suitably learn the patterns relevant biological problem to which they have been applied and that the descriptors contirbuting to these are sensible and that the algorithm (random forest) is not learning spurious or artificial patterns within the data (ref).
Approaches such as LIME and model response surfaces enable investigators to query the contributions that phenotypic descriptors have in particular cases of interest.
This inturn promotes trust in the indexes derived from these models and reduces their black box nature (ref).

### Applications

- Explain the wider application of the approach

@@

This would enable the development of web based tools that can then provide access to

- Discussion of applications to forestry monitoring and management
- Refinement of protocol for large scale surveying or alternative applications

@@

The current context in which these decline indexes are presented is based on a very focus set of descriptors on sites with a high proportion of declining trees. 
Scaling this technique to enable generalised naive computation of decline indexes for any tree sampled in the UK would require a substantially larger training data set that incorporates the descriptive diversity of oak decline in the population as whole (ref).
Scaling to this level would also require an appropriate method of tree size correction that could be applied on an individual tree basis.
A potential solution to this would be to correct trees based on estimated age from DBH as proposed by @white1998.

@@

- Crown condition descriptors most important for describing overall decline severity in the PDI.
- Bacterial bleed related descriptors are most important for describing decline acuteness.
- Decline indexes provide intuitive measures of decline severity based on simple phenotypic measurements that do not require specialist equipment

The decline indexes presented here provided intuitive measures 
The approach presented here is easily adaptable to incorporate any phenotype
It is encouraged that this approach of deriving descriptive index measurse in other tree health phenotyping contexts.
